\thetitle

### 2008-09-02 paper/paper.tex
paper: Many improvements to all sections except the introduction and the sections dealing with the proof.
darcs-hash:20080902214607-bab43-eedf50bb80530cc184d292794f423b124d507c2a.gz

% todo:
% - check that names are spelled and encoded right in bibliography
% - lhs2tex shows parameters declared with {...} differently from those declared with (...)
% - mention that we don't do correctness proofs (although we can add them in the formalization, of course)

@@

The formalization is based on the ``paper proof'' in \cite{introtoalgos}, which consists of three parts. The first part argues that the total number of comparisons performed by the algorithm can be written as a sum of probabilities of comparisons between individual pairs of elements from the input list. The second part then derives from the algorithm a specific formula for those probabilities. Finally, the third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-probabilities. Thus, the first part essentially establishes an elementwise-comparison-probability lemma as an interface between the second and third part, proved by the former and used by the latter.

@@

The embedding is introduced in section \ref{embed}, where we demonstrate its use by first defining a simple deterministic Quicksort implementation (which just selects the head of the input list as its pivot element, and uses two separate filter passes to partition the input list), and then proving that implementation's quadratic worst-case complexity.

@@

The next two sections detail the actual formalized proof. Section \ref{reduction} corresponds to the first part in the original proof, and shows how the main theorem can be split up into a lemma (stated in terms of another special-purpose monad) about the probability of comparisons being performed for pairs of input elements indexed a certain way, and a strictly numerical part.

@@

Section \ref{finishing} shows the proof of the lemma mentioned above, and corresponds to the part in the original proof that derives the probability formula from the algorithm. Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason about the detailed behavior of the algorithm, it is by far the most involved part of the formalization.

@@

Before the main theorem can even be formally stated, let alone proved, a formal definition of the algorithm is needed, expressed in a programming language with well-defined semantics, suitable for subsequent formal reasoning. Here, we exploit the computational nature of the type theory on which Coq is based. Rather than defining the syntax and semantics of a programming language from scratch (this would be a ``deep'' embedding) like we would have to do in, say, ZFC mathematics, we instead use the built-in lambda calculus as a functional programming language, and write the algorithm as a specially adorned lambda term. More specifically, we express the algorithm monadically, with both the monad and the counted operation (in our case, the elementwise comparison---the usual complexity metric for sorting algorithms) made into parameters. Expressed this way, a deterministic quicksort that simply selects the head of the input list as its pivot element, and uses two simple filter passes to partition the input list, looks as follows:

@@

  Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
    match l with
    | nil => ret nil
    | pivot :: t =>
        lower <- filter (gt pivot) t >>= qs;
        upper <- filter (le pivot) t >>= qs;
        ret (lower ++ pivot :: upper)
    end.
\end{code}
|Monad| is a dependent record containing the (coercible) carrier of type |Set -> Set|, along with |bind| (infix: |>>=|) and |ret| (for ``return'') operations, and the three monad laws. For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}. The notation |x <- y; z|, shorthand for |y >>= \x -> z| (where |x| may occur in |z|) is known as ``do-notation''.

@@

For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer called |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad.

@@

We use |MMT| to piggyback the additive monoid on |nat| onto the identity monad, and lift elementwise comparison into the resulting monad, which we call |SP| (for ``simply-profiled'').
\begin{code}
  Definition SP: Monad := MMT NatAddMonoid IdMonad.

@@

\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\end{propo}
  \todo{get rid of period after ``Proposition'' above}
\begin{proof}
  The proof is by induction matching |qs|'s recursion. For an empty input list, we have |cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|. For a non-empty input list |(pivot :: t)|, the cost decomposes into
  \begin{code}
    cost (filter (le pivot) t) + cost (qs_SP (result (filter (lt pivot) t))) +
    cost (filter (gt pivot) t) + cost (qs_SP (result (filter (gt pivot) t))) +
    cost (ret (result (qs_SP (result (filter (lt pivot) t))) ++
     pivot :: result (qs_SP (result (filter (gt pivot) t))))).
  \end{code}
  The |filter| costs are easily proved (by induction on |t|) to be |length t| each. The cost of the final |ret| is 0 by definition. The induction hypothesis applies to the recursive |qs_SP| calls. Furthermore,
  \begin{code}
  length (result (filter (le pivot) t)) +
  length (result (filter (gt pivot) t)) <= length t
  \end{code}
  can easily be proved by induction on |t|, because the two predicates filtered on are mutually exclusive. Abstracting |filter (le pivot) t| and its |gt| counterpart, this leaves
  \begin{code}
    forall (t flt flt': list T), length flt + length flt' <= length t ->
      length t + sqrd (length flt) + length t + sqrd (length flt')  + 0 <= sqrd (S (length t)),
  \end{code}
  which is true by elementary arithmetic. \qed
\end{proof}

We now expand the machinery in preparation of the average-case complexity proof.

@@

The Quicksort algorithm used by the average-case proof we formalize differs from the one presented in the last section in two ways: it is nondeterministic, and uses a single three-way partition pass instead of two two-way filter passes. Combined, these two traits ensure that the $O(n \log n)$ average-case property holds not just averaged over all input lists, but for each individual input list as well; nondeterministic pivot selection avoids the pathological cases any deterministic pivot selection strategy inevitably suffers, while a single three-way partition pass avoids the pathological quadratic case that two two-way filter passes suffer for an input list consisting of elements all of the same value. This greatly simplifies things, because it means that the global bound will follow immediately if it is proved for an arbitrary input.

Hence, we define

@@

  Fixpoint partition (pivot: T) (l: list T): M (Partitioning T) :=
    match l with
    | nil => ret emptyPartitioning
    | h :: t =>
        b <- cmp h pivot;
        tt <- partition pivot t ;
        ret (addToPartitioning b h tt)
    end.

  Variable pick: forall (A: Set), ne_list A -> M A.

  Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
    match l with
    | nil => ret nil
    | _ =>
        i <- pick [0 ... length l - 1];
        let pivot := nth l i in
        part <- partition pivot (remove l i);
        low <- qs (part Lt);
        upp <- qs (part Gt);
        ret (low ++ pivot :: part Eq ++ upp)
    end.
\end{code}}
Here, |comparison| is an enumeration with values |Lt|, |Eq|, and |Gt|. A |Partitioning T| is a function of type |comparison -> list T|. An |ne_list T| is a non-empty list of |T|'s. The functions |nth| and |remove| select and remove the $n$th element of a list, respectively.

@@

  Definition pick_ne_tree (T: Set): ne_list T -> M_ne_tree T
    := Node fcmp ne_list.map Leaf.
\end{code}
(We use non-empty trees because we do not consider computations that produce no values at all, and using potentially empty trees would complicate the definition of a tree's average value below.)

@@

Because we are often interested in the expected value of a measure of the output of a program that does not return numbers, we further define

@@

To form the monad with which we will instantiate |qs| for the main theorem, we now piggyback the additive monoid on |nat| onto |M_ne_tree| using |MMT|, and call the result |NDP| (for ``non-deterministically profiled''):
\begin{code}
  Definition M_NDP: Monad := MMT NatAddMonoid M_ne_tree.

@@

  Definition qs_NDP := qs M_NDP le_NDP (lift pick_ne_tree).
\end{code}

When working with computations in a monad formed by transforming |M_ne_tree| using |MMT|, we are often interested in the expected value of a function of the piggybacked monoid. For this, we introduce |monoid_expec|.
\begin{code}
  Definition monoid_expec (m: Monoid) (f: m -> nat) {A: Set}
    : (MonoidMonadTrans.M m NeTreeMonad A) -> R := expec (f fcmp fst).
\end{code}
Since |cost = fst|, we have |forall t, expec cost t = monoid_expec id t|.

In addition to all the identities |monoid_expec| inherits from |expec|, it has some of its own. One identity worth highlighting states that if one transforms |M_ne_tree| using a monoid |m|, then for a monoid homomorphism |h| from |m| to the additive monoid on |nat|, |monoid_expec h| distributes over |bind|, provided that the expected monoid value of the right hand side does not depend on the computed value of the left hand side:
\begin{code}
  monoid_expec_plus: forall (m: Monoid) (h: m -> nat) (A B: Set)
    (f: MMT m NeTreeMonad A) (g: A -> MMT m NeTreeMonad B):
    (forall x y `In` f -> monoid_expec h (g (snd x)) = monoid_expec h (g (snd y))),
    monoid_expec h (f >>= g) =
      monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))).
\end{code}
Since |id| is a monoid homomorphism, |monoid_expec_plus| applies to |NDP| and |expec cost|. In section \ref{reduction}, we will use it with another monoid and homomorphism.

@@

Now that we have a solid definition of nondeterministic, instrumented quicksort, as well as the means to express the expected values of nondeterministic programs, the last thing needed before the main theorem can be stated, is the notion of big-O complexity. The definition we use is a standard textbook definition, except that we work with a separate measure function parameter:

@@

  Notation {-"``"-}over m ,{-"\ "-} f =O( g ){-"\text{''''}"-} := (measured_bigO m f g).
\end{code}}
We now state the main theorem.
\begin{code}
  Theorem qs_avg: over length,
    expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
\end{code}
Thanks to the property discussed at the start of the previous section, the above follows as a corollary from the (slightly stronger) statement that
\begin{code}
  Theorem qs_expec_cost:
    forall l, expec cost (qs_NDP l) <= 2 * length l * S (log2ceil (length l)),
\end{code}
the proof of which is described in the remainder of this paper.

@@

\begin{itemize}
\item We have pretended to have used ordinary natural numbers as indices into ordinary lists, completely ignoring the inescapable issues of index validity that could not be ignored in the actual formalization. There, we use vectors (lists whose size is part of their type) and bounded natural numbers in many places instead. Using these substantially reduces the amount of |i < length l| proofs that need to be produced, converted, and passed around, but the issue is still far from painless.
\item Using the |Program| facility to deal with quicksort's non-structural recursion was not completely as trivial as we made it out to be. To make proving the generated proof obligations possible, the types of |filter| and |partition| had to be sigma-decorated with modest length guarantees, because the opacity of the unspecified monad's |bind| operation prevented |Program| from being able to ``see'' the relation between the function's parameters and the arguments it passed in its recursive calls.
\item Using the induction principles corresponding to the definitions produced by |Program| for a function already defined with two layers of monadicity quickly became unwieldy. To deal with this, we defined a series of specialized induction principles for specific instances of |qs|, which put the recursive proof obligation in as simple terms as possible, with as little monadic indirection as possible, using among other things the observations regarding determinism of the comparison operation, as described in section \ref{finishing}.
\end{itemize}

@@

In conclusion, we have found the shallow monadic embedding to be a straightforward, clean, and solid basis for reasoning about an algorithm's computational complexity. Calling it an embedding almost seems misplaced, since the algorithm definitions are essentially just monadically expressed functional programs in Coq's calculus. Furthermore, since monads and monad transformers have been used in functional programming for decades, the basic idea should be immediately clear to any functional programmer. Still, it is worth noting that this pleasing convergence of subject and object languages is due solely to the fact that Coq is based on type theory.

Quicksort turned out to be a good stress test for the technique, showing that it copes well with nondeterminism and nontrivial average-case bounds. The |U| monad shows that custom profiling monads can be defined to analyse very algorithm-specific traits for a specific part of a larger complexity proof, which was crucial for this particular proof.

For future work, a logical next challenge is to try and formalize a (likely more complex) proof that the $O(n \log n)$ bound holds also for deterministic quicksorts (with sensible pivot selection strategies).

>>>>

% todo:
% - check that names are spelled and encoded right in bibliography
% - lhs2tex shows parameters declared with {...} differently from those declared with (...)
% - mention that we don't do correctness proofs (although we can add them in the formalization, of course)
% - decide whether to hide, ignore, or explain R/N differences. explaining costs precious space. ignoring is "begging the question". hiding is naughty.
% - have a sentence somewhere hinting at the complications caused by duplicate elements (e.g., by pointing out simplifications that hold only with the assumption of no duplicate elements).

@@

The formalization is based on the ``paper proof'' in \cite{introtoalgos}, which consists of three parts. The first part argues that the total number of comparisons performed by the algorithm can be written as the sum of the expected number of comparisons between individual pairs of elements from the input list (ordered a certain way). The second part then derives from the algorithm a specific formula for this expectancy. Finally, the third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-expectancies. Thus, the first part essentially establishes a lemma about the expected number of comparisons between individual pairs of elements from the input list as an interface between the second and third part, proved by the former and used by the latter.

@@

The embedding is introduced in section \ref{embed}, where we demonstrate its use by first defining a simple deterministic Quicksort implementation (which just selects the head of the input list as its pivot element, and uses two separate filter passes to partition the input list), and then defining an operation counting monad that lets us prove that implementation's quadratic worst-case complexity.

@@

The next two sections detail the actual formalized proof. Section \ref{reduction} corresponds to the first part in the original proof, and shows how the main theorem can be split up into a lemma (stated in terms of another special-purpose monad) about the expected number of comparisons being performed for pairs of input elements indexed a certain way, and a strictly numerical part.

@@

Section \ref{finishing} shows the proof of the lemma mentioned above, and corresponds to the part in the original proof that derives the expectancy formula from the algorithm. Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason in detail about the behavior of the algorithm, it is by far the most involved part of the formalization.

@@

% todo: less pompous here \/

As stated before, the key to giving a proper formal treatment of those parts of the proof for which the original contents itself with appeals to intuition, lies in the use of an appropriate representation of the algorithm. Indeed, we cannot even formally state the main theorem until we have both an algorithm definition and the means to denote its use of comparisons (the usual complexity metric for sorting algorithms).

Since we are working in Coq, we already have at our disposal a full functional programming language, in the form of Coq's pCIC \cite{coq'art}. However, just writing the algorithm as an ordinary Coq function would not let us observe its use of comparisons. We make those uses observable by writing the algorithm as a monadically expressed function, parameterizing it on both the monad itself and on the comparison operation. Expressed this way, a deterministic quicksort that simply selects the head of the input list as its pivot element, and uses two simple filter passes to partition the input list, looks as follows:

@@

  Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
    match l with
    | nil => ret nil
    | pivot :: t =>
        lower <- filter (gt pivot) t >>= qs;
        upper <- filter (le pivot) t >>= qs;
        ret (lower ++ pivot :: upper)
    end.
\end{code}
|Monad| is a dependent record containing the (coercible) carrier of type |Set -> Set|, along with |bind| (infix: |>>=|) and |ret| (for ``return'') operations, and the three monad laws. For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}. The notation ``|x <- y; z|'', shorthand for |y >>= \x => z| (where |x| may occur in |z|) is known as ``do-notation''.

@@

For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer called |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.

@@

We now use |MMT| to piggyback the additive monoid on |nat| onto the identity monad, and lift elementwise comparison into the resulting monad, which we call |SP| (for ``simply-profiled'').
\begin{code}
  Definition SP: Monad := MMT NatAddMonoid IdMonad.

@@

\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\footnote{(We don't use big-$O$ notation for this simple statement, as it would only obfuscate. Big-$O$ complexity is discussed in section \ref{statement}.)}
\end{propo}

\begin{proof}
  The proof is by induction matching |qs|'s recursion. For an empty input list, we have |cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|. For a non-empty input list |(pivot :: t)|, the cost decomposes into
  \begin{code}
    cost (filter (le pivot) t) + cost (qs_SP (result (filter (lt pivot) t))) +
    cost (filter (gt pivot) t) + cost (qs_SP (result (filter (gt pivot) t))) +
    cost (ret (result (qs_SP (result (filter (lt pivot) t))) ++
     pivot :: result (qs_SP (result (filter (gt pivot) t))))).
  \end{code}
  The |filter| costs are easily proved (by induction on |t|) to be |length t| each. The cost of the final |ret| is 0 by definition. The induction hypothesis applies to the recursive |qs_SP| calls. Furthermore, by induction on |t|, we can easily prove
  \begin{code}
  length (result (filter (le pivot) t)) +
  length (result (filter (gt pivot) t)) <= length t,
  \end{code}
  because the two predicates filtered on are mutually exclusive. Abstracting |filter (le pivot) t| and its |gt| counterpart, this leaves
  \begin{code}
    forall (t flt flt': list T), length flt + length flt' <= length t ->
      length t + sqrd (length flt) + length t + sqrd (length flt') + 0 <= sqrd (S (length t)),
  \end{code}
  which is true by elementary arithmetic. \qed
\end{proof}

We now expand the technique in preparation of the average-case complexity proof.

@@

The Quicksort algorithm used by the average-case proof we formalize differs from the one presented in the last section in two ways. First, it selects a pivot element nondeterministically, thereby avoiding the pathological quadratic cases any deterministic pivot selection strategy inevitably suffers. Second, it uses a single three-way partition pass instead of two two-way filter passes, to avoid the pathological quadratic case that two two-way filter passes suffer for an input list consisting of elements all of the same value.

Combined, these two traits ensure that the $O(n \log n)$ average-case bound holds not just averaged over all input lists, but for each individual input list as well. This greatly simplifies things, because it means that the global bound can be proved simply by proving it as a bound of the expected number of comparisons performed for an arbitrary input.

Hence, we now modify our |qs| definition to use a three-way partition pass, and to use nondeterministic pivot selection. Following that, we will compose a new monad (again using |MMT|, but this time transforming a nondeterminism monad) that the new definition can be instantiated with to capture its \emph{expected} number of comparisons.

First, we define |partition|. Instead of a two-way comparison operation producing a monadic |bool|, it takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|. |partition| constructs a |Partitioning T|, which is simply a function of type |comparison -> list T|. We might equally well have used an inductive definition of |Partitioning T| as a record containing three lists. % at least until i remember why i used a function.

@@

  Fixpoint partition (pivot: T) (l: list T): M (Partitioning T) :=
    match l with
    | nil => ret emptyPartitioning
    | h :: t =>
        b <- cmp h pivot;
        tt <- partition pivot t ;
        ret (addToPartitioning b h tt)
    end.
\end{code}}

Next, we redefine |qs| to use |partition|, and have it take as an additional parameter a |pick| operation, representing nondeterministic selection of an element of a non-empty list of choices. (An |ne_list T| is a non-empty list of |T|'s.)

\begin{code}
  Variable pick: forall (A: Set), ne_list A -> M A.

  Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
    match l with
    | nil => ret nil
    | _ =>
        i <- pick [0 ... length l - 1];
        let pivot := nth l i in
        part <- partition pivot (remove l i);
        low <- qs (part Lt);
        upp <- qs (part Gt);
        ret (low ++ pivot :: part Eq ++ upp)
    end.
\end{code}
The functions |nth| and |remove| select and remove the $n$th element of a list, respectively.

@@

  Definition pick_ne_tree (T: Set): ne_list T -> M_ne_tree T
    := Node fcmp ne_list.map Leaf.
\end{code}
We use non-empty trees because we do not consider partial functions, and using potentially empty trees would complicate the definition of a tree's average value below. This is also why we used |ne_list| for |pick|.

@@

To denote the expected value of a \emph{measure} of the output of a program, we define

@@

To form the monad with which we will instantiate |qs| for the main theorem, we now piggyback the additive monoid on |nat| onto |M_ne_tree| using |MMT|, and call the result |NDP| (for ``nondeterministically profiled''):
\begin{code}
  Definition M_NDP: Monad := MMT NatAddMonoid M_ne_tree.

@@

  Definition qs_NDP := qs M_NDP le_NDP (lift pick_ne_tree).
\end{code}
We can now denote the expected number of comparisons of a |qs_NDP| application by |expec cost (qs_NDP l)|, and will use this in our statement of the main theorem in the next section.

But before we do so, we define a slight refinement of |expec| that specifically observes the monoid component of computations in monads formed by transforming |M_ne_tree| using |MMT| (like |NDP|).
\begin{code}
  Definition monoid_expec (m: Monoid) (f: m -> nat) {A: Set}
    : (MonoidMonadTrans.M m NeTreeMonad A) -> R := expec (f fcmp fst).
\end{code}
Since |cost = fst|, we have |forall t, expec cost t = monoid_expec id t|.

In addition to all the identities |monoid_expec| inherits from |expec|, it has some of its own. One identity states that if one transforms |M_ne_tree| using a monoid |m|, then for a monoid homomorphism |h| from |m| to the additive monoid on |nat|, |monoid_expec h| distributes over |bind|, provided that the expected monoid value of the right hand side does not depend on the computed value of the left hand side:

\parbox{\textwidth}{\begin{code}
  monoid_expec_plus: forall (m: Monoid) (h: m -> nat) (A B: Set)
    (f: MMT m NeTreeMonad A) (g: A -> MMT m NeTreeMonad B):
    (forall x y `In` f -> monoid_expec h (g (snd x)) = monoid_expec h (g (snd y))),
    monoid_expec h (f >>= g) =
      monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))).
\end{code}}

Since |id| is a monoid homomorphism, |monoid_expec_plus| applies to |NDP| and |expec cost|. In section \ref{reduction}, we will use |monoid_expec_plus| with another monoid and homomorphism.

@@

The last thing needed before the main theorem can be stated, is the notion of big-O complexity. The definition we use is a standard textbook definition, except that we work with a separate measure function parameter:

@@

  Notation {-"``"-}over m ,{-"\ "-} f =O( g ){-"\text{''''}"-} := (measured_bigO m f g).
\end{code}}
We now state the main theorem.
\begin{code}
  Theorem qs_avg: over length,
    expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
\end{code}
Thanks to the property discussed at the start of the previous section, % perhaps too far back
the above follows as a corollary from the (slightly stronger) statement that
\begin{code}
  Theorem qs_expec_cost:
    forall l, expec cost (qs_NDP l) <= 2 * length l * S (log2ceil (length l)),
\end{code}
the proof of which is described in the remainder of this paper.

@@

\begin{itemize}
\item We have pretended to have used ordinary natural numbers as indices into ordinary lists, completely ignoring the inescapable issues of index validity that could not be ignored in the actual formalization. There, we use vectors (lists whose size is part of their type) and bounded natural numbers in many places instead. Using these substantially reduces the amount of |i < length l| proofs that need to be produced, converted, and passed around, but this solution is still far from painless.
\item Using the |Program| facility to deal with quicksort's non-structural recursion was not completely as trivial as we made it out to be. To make proving the generated proof obligations possible, the types of |filter| and |partition| had to be sigma-decorated with modest length guarantees, because the opacity of the unspecified monad's |bind| operation prevented |Program| from being able to ``see'' the relation between the function's parameters and the arguments it passed in its recursive calls.
\item Using the induction principles corresponding to the definitions produced by |Program| for a function already defined with two layers of monadicity quickly became unwieldy. To deal with this, we defined a series of specialized induction principles for specific instances of |qs|, which put the recursive proof obligation in as simple terms as possible, with as little monadic indirection as possible, using among other things the observations regarding determinism of the comparison operation, as described in section \ref{finishing}.
\end{itemize}

@@

The shallow monadic embedding provides a simple but effective representation of the algorithm. The parameterization of the monad makes the representation very flexible, allowing a single definition to be instantiated either with basic monads (like the identity monad or bare nondeterminism monads) to get a non-instrumented version suitable for extraction and correctness proofs, or with monads that use |MMT| to add operation-counting, to support our complexity proofs.

Furthermore, since the shallowness of the embedding lets us re-use all standard Coq data types and facilities, including the powerful |Program Fixpoint| command, the actual algorithm definition itself is reasonably clean.

The |MMT| monoid monad transformer provided the ``glue'' to add instrumentation onto a base monad. Its use of a monoid to represent the operation ``counter'' let us use a single set of lemmas (about |monoid_expec|) for |NDP| and |U|, as both were instances of the |M_ne_tree| monad transformed with |MMT|, differing only in the monoid used. Indeed, the relative ease with which we could transform the expectancy lemma into the |U| monad, a monad highly specialized for the specific order-index based counting scheme we used, further attests to the flexibility of the combination of a monadic representation and the |MMT| monad transformer.

\todo{Paragraph saying that the formalized proof is faithful to the original proof, except for the part where we dropped the duplicate elements assumption. The purely mathematical part could be directly transcribed. The only theory we had to add served to support a formal treatment of the parts that the original clearly took for granted.}

% old passage that may yet be useful in some form or another:
%    Calling it an embedding almost seems misplaced, since the algorithm definitions are essentially just monadically expressed functional programs in Coq's calculus. Furthermore, since monads and monad transformers have been used in functional programming for decades, the basic idea should be immediately clear to any functional programmer. Still, it is worth noting that this pleasing convergence of subject and object languages is due solely to the fact that Coq is based on type theory.

For future work, an interesting next challenge is to try and formalize a (likely more complex) proof that the $O(n \log n)$ bound holds also for deterministic quicksorts (with sensible pivot selection strategies).
